---
title: "Data Wrangling in R"
author: "Ryan Musasia"
filters:
  - webr
webr:
  packages:
    - tidyverse
    - sf
  resources:
    - ideal3a.csv
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: false
    logo: cema_logo.png
    progress: true
    transition: fade
    transition-speed: fast
    highlight-style: github
    code-copy: true
    code-block-bg: true
    code-line-numbers: false
    width: 1280
    height: 720
    margin: 0.09
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
---

```{=html}
<style>
/* ── Option 2: Two-Tier Typography ── */
/* Prose — prominent and readable */
.reveal .slides section p,
.reveal .slides section li {
  font-size: 22px !important;
  line-height: 1.6 !important;
}
/* Blockquotes — one step down from prose */
.reveal .slides section blockquote p,
.reveal .slides section blockquote li {
  font-size: 17px !important;
  line-height: 1.55 !important;
}
/* Slide tables — compact but legible */
.reveal .slides section table td,
.reveal .slides section table th {
  font-size: 15px !important;
  line-height: 1.5 !important;
  padding: 5px 10px !important;
}
/* WebR editor code — clearly smaller than prose */
.cm-editor,
.cm-content,
.cm-line {
  font-size: 14px !important;
  line-height: 1.55 !important;
}
/* Static {r} eval:false code blocks */
.reveal div.sourceCode pre,
.reveal div.sourceCode pre code {
  font-size: 14px !important;
  line-height: 1.55 !important;
}
/* Output — smallest tier (text output) */
pre:not([class]) {
  background: #faf9f7 !important;
  border: none !important;
  border-radius: 0 !important;
  padding: 10px 14px !important;
  font-size: 11px !important;
  color: #3D3D3D !important;
  line-height: 1.5 !important;
  margin: 0 !important;
  overflow-y: auto !important;
  max-height: 190px !important;
}
/* ── Plot output from WebR ── */
.qwebr-output-code-execution canvas,
.qwebr-output-code-execution img {
  max-width: 100% !important;
  height: auto !important;
  border-radius: 4px !important;
  margin-top: 6px !important;
}

/* ── Dual logo: same height, blended, side by side ── */
.reveal .slide-logo {
  bottom: 22px !important;
  right: 155px !important;
  max-height: 44px !important;
}
.reveal .slide-logo img {
  max-height: 44px !important;
  width: auto !important;
}
.reveal::after {
  content: "";
  display: block;
  position: fixed;
  bottom: 8px;
  right: 10px;
  width: 135px;
  height: 90px;
  background: url("NPHI.png.jpeg") no-repeat center / contain;
  mix-blend-mode: multiply;
  z-index: 999;
  pointer-events: none;
}

/* ── Button fixes ── */
div.sourceCode {
  position: relative !important;
}
.code-copy-button {
  position: absolute !important;
  top: 6px !important;
  right: 6px !important;
  z-index: 999 !important;
  pointer-events: all !important;
}
.qwebr-button-reset,
.qwebr-button-copy {
  pointer-events: all !important;
  position: relative !important;
  z-index: 999 !important;
}
</style>
<script>
(function () {
  /* ── Section tracker ── */
  var NUM_SECTIONS = 9;
  function buildTracker() {
    if (document.getElementById("section-tracker")) return;
    var el = document.createElement("div");
    el.id = "section-tracker";
    var pipe = document.createElement("div");
    pipe.className = "tracker-pipe";
    el.appendChild(pipe);
    for (var i = 1; i <= NUM_SECTIONS; i++) {
      var sp = document.createElement("span");
      sp.textContent = i;
      el.appendChild(sp);
    }
    document.body.appendChild(el);
  }
  function updateTracker() {
    var current = Reveal.getCurrentSlide();
    if (!current) return;
    var allTop = Array.from(document.querySelectorAll(".reveal .slides > section"));
    var currentTop = current.closest(".reveal .slides > section");
    var idx = allTop.indexOf(currentTop);
    document.querySelectorAll("#section-tracker span").forEach(function(sp, i) {
      sp.classList.toggle("active", i + 1 === idx);
    });
    var tracker = document.getElementById("section-tracker");
    if (tracker) tracker.style.display = idx === 0 ? "none" : "flex";
  }
  document.addEventListener("DOMContentLoaded", function () {
    buildTracker();
    if (typeof Reveal !== "undefined") {
      Reveal.on("ready", function() { buildTracker(); updateTracker(); });
      Reveal.on("slidechanged", updateTracker);
    }
  });
  setTimeout(buildTracker, 600);
  /* ── Copy & Reset button feedback ── */
  document.addEventListener("click", function (e) {
    var copyBtn = e.target.closest(".code-copy-button, .qwebr-button-copy");
    var resetBtn = e.target.closest(".qwebr-button-reset");
    if (copyBtn) {
      showFeedback(copyBtn, "✓ Copied!", "#639922");
    }
    if (resetBtn) {
      showFeedback(resetBtn, "↺ Reset!", "#B22222");
    }
  });
  function showFeedback(btn, msg, color) {
    var prev = btn.innerHTML;
    var prevBg = btn.style.background;
    var prevColor = btn.style.color;
    var prevBorder = btn.style.border;
    btn.textContent = msg;
    btn.style.background = color;
    btn.style.color = "#fff";
    btn.style.border = "none";
    setTimeout(function () {
      btn.innerHTML = prev;
      btn.style.background = prevBg;
      btn.style.color = prevColor;
      btn.style.border = prevBorder;
    }, 1500);
  }
  /* ── Output card wrapper (text output only) ── */
  document.addEventListener("DOMContentLoaded", function () {
    var observer = new MutationObserver(function () {
      document.querySelectorAll("pre:not([class])").forEach(function (pre) {
        if (pre.parentNode.classList.contains("out-card-wrap")) return;
        var wrap = document.createElement("div");
        wrap.className = "out-card-wrap";
        wrap.style.cssText = "background:#faf9f7;border:0.5px solid #d1cab8;border-left:3px solid #639922;border-radius:6px;overflow:hidden;margin-top:8px;max-height:220px;display:flex;flex-direction:column;";
        var header = document.createElement("div");
        header.style.cssText = "background:#ede9e1;padding:5px 12px;font-size:10px;color:#7a6f60;font-weight:500;display:flex;align-items:center;gap:6px;flex-shrink:0;border-bottom:0.5px solid #d1cab8;";
        var meta = '';
        var tibbleMatch = pre.textContent.match(/#\s*A tibble:\s*([\d,]+)\s*[×x]\s*(\d+)/);
        if (tibbleMatch) {
          meta = '<span style="font-size:9px;background:#e8e2d8;color:#7a6f60;padding:1px 6px;border-radius:3px;font-family:monospace;margin-left:6px;">' + tibbleMatch[1] + ' rows × ' + tibbleMatch[2] + ' cols</span>';
        }
        header.innerHTML = '<span style="width:6px;height:6px;background:#639922;border-radius:50%;display:inline-block;"></span> Output' + meta + '<span style="color:#bbb;font-size:9px;margin-left:auto;">scroll ↕</span>';
        pre.parentNode.insertBefore(wrap, pre);
        wrap.appendChild(header);
        wrap.appendChild(pre);
      });
    });
    observer.observe(document.body, { childList: true, subtree: true });
    /* ── Suppress WebR output on code-only slides (chart shown on next slide) ── */
    function hideCodeOnlyOutputs() {
      document.querySelectorAll(".code-only .qwebr-output-code-execution, .code-only .out-card-wrap").forEach(function (el) {
        el.style.setProperty("display", "none", "important");
      });
    }
    /* Watch childList AND attribute changes — webr reveals output by toggling style, not inserting nodes */
    var codeOnlyObs = new MutationObserver(hideCodeOnlyOutputs);
    codeOnlyObs.observe(document.body, {
      childList: true,
      subtree: true,
      attributes: true,
      attributeFilter: ["style", "class"]
    });
    /* Belt-and-suspenders: poll for the first 60 s to catch any remaining races */
    var _coTimer = setInterval(hideCodeOnlyOutputs, 300);
    setTimeout(function () { clearInterval(_coTimer); }, 60000);
    /* Also fire on Reveal slide transitions */
    setTimeout(function () {
      if (typeof Reveal !== "undefined") {
        Reveal.on("ready", hideCodeOnlyOutputs);
        Reveal.on("slidechanged", hideCodeOnlyOutputs);
      }
      hideCodeOnlyOutputs();
    }, 800);
  });
})();
</script>
```

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(readxl)
if (!requireNamespace("sf", quietly = TRUE)) install.packages("sf")
library(sf)
ideal      <- read_csv("https://raw.githubusercontent.com/ThumbiMwangi/R_sources/master/ideal3a.csv")
population <- read_csv("https://raw.githubusercontent.com/Musasia-Ryan/Introduction-to-R/main/aggregated_Population.csv")
cattle     <- population |> filter(Species == "Cattle(Total)")
kenya <- tryCatch(
  read_sf("https://raw.githubusercontent.com/Musasia-Ryan/Introduction-to-R/main/ken_county.gpkg"),
  error = function(e) { message("ken_county.gpkg not found — map slides will be blank"); NULL }
)
sublocation_info <- tibble(
  sublocation = c("East Siboti", "Kidera", "Kokare", "Kamunuoit", "Karisa",
                  "Igero", "Bukati", "Bumala A", "Namboboto", "Ojwando B",
                  "Luanda", "Bujwanga", "Magombe East"),
  division    = c("Sirisia", "Sirisia", "Sirisia", "Teso", "Teso",
                  "Butula", "Butula", "Butula", "Nambale", "Nambale",
                  "Emuhaya", "Emuhaya", "Emuhaya"),
  altitude_m  = c(1450, 1380, 1520, 1340, 1290,
                  1180, 1220, 1250, 1350, 1290,
                  1420, 1380, 1310)
)
```


```{webr-r}
#| label: webr-data-load
#| autorun: true
#| include: false
download.file(
  "https://raw.githubusercontent.com/Musasia-Ryan/Introduction-to-R/main/ideal3a.csv",
  "ideal3a.csv"
)
download.file(
  "https://raw.githubusercontent.com/Musasia-Ryan/Introduction-to-R/main/aggregated_Population.csv",
  "aggregated_Population.csv"
)
download.file(
  "https://raw.githubusercontent.com/Musasia-Ryan/Introduction-to-R/main/ken_county.gpkg",
  "ken_county.gpkg"
)
library(tidyverse)
library(sf)
ideal <- read_csv("ideal3a.csv")
population <- read_csv("aggregated_Population.csv")
cattle <- population |> filter(Species == "Cattle(Total)")
kenya <- read_sf("ken_county.gpkg")
sublocation_info <- tibble(
  sublocation = c("East Siboti", "Kidera", "Kokare", "Kamunuoit", "Karisa",
                  "Igero", "Bukati", "Bumala A", "Namboboto", "Ojwando B",
                  "Luanda", "Bujwanga", "Magombe East"),
  division    = c("Sirisia", "Sirisia", "Sirisia", "Teso", "Teso",
                  "Butula", "Butula", "Butula", "Nambale", "Nambale",
                  "Emuhaya", "Emuhaya", "Emuhaya"),
  altitude_m  = c(1450, 1380, 1520, 1340, 1290,
                  1180, 1220, 1250, 1350, 1290,
                  1420, 1380, 1310)
)
# Pre-compute visualisation datasets so plot slides work on direct navigation
calves_per_sub <- ideal |>
  group_by(sublocation) |>
  summarise(n_calves = n_distinct(CalfID)) |>
  ungroup()
calves_by_sex <- ideal |>
  mutate(sex_label = case_when(
    CalfSex == 1 ~ "Male",
    CalfSex == 2 ~ "Female",
    TRUE ~ "Unknown"
  )) |>
  group_by(sublocation, sex_label) |>
  summarise(n_calves = n_distinct(CalfID), .groups = "drop")
```

## Today's Session

:::: {.columns}
::: {.column width="50%"}
**Getting data in & oriented**

- Reading data — `read_csv()`, `read_excel()` <br><span class="muted small">flat files and spreadsheets</span>
- Inspecting — `glimpse()`, `summary()` <br><span class="muted small">know your data before you touch it</span>
- The pipe `|>` <br><span class="muted small">chain steps into readable sentences</span>
- Rows — `filter()`, `arrange()` <br><span class="muted small">ask questions of your data</span>
:::
::: {.column width="50%"}
**Shaping & summarising**

- Columns — `select()`, `rename()` <br><span class="muted small">keep only what matters</span>
- Deriving variables — `mutate()`, `case_when()` <br><span class="muted small">create meaning from raw values</span>
- Aggregation — `group_by()` + `summarise()` <br><span class="muted small">collapse many rows into insight</span>
- Joining tables <br><span class="muted small">unite data from multiple sources</span>

**Visualising**

- `ggplot2` — grammar of graphics <br><span class="muted small">bar charts, column charts, maps</span>
:::
::::

# Reading Data into R

Before you can analyse data, you must load it into R. Two tidyverse packages handle virtually every file format you will encounter in practice.

## Key Packages

- `readr` reads rectangular text files.
- `readxl` reads Excel workbooks.
- Both produce a **tibble** — R's modern, opinionated data frame.

| Package | Function | Use when… |
|---------|----------|-----------|
| `readr` | `read_csv()` | Your file is `.csv` — the universal format |
| `readr` | `read_tsv()` | Columns are separated by tabs |
| `readxl` | `read_excel()` | Your file is `.xlsx` or `.xls` |

```{r}
#| label: load-libs
#| eval: false
library(tidyverse)
library(readxl)
```

> `readxl` is not part of the tidyverse core. Always call `library(readxl)` explicitly.

## Reading a CSV File

- `read_csv()` is the workhorse for flat files. It reads faster than base R's `read.csv()`, it never silently converts strings to factors, and prints a column-type summary on load so you always know what was parsed.

```{r}
#| label: read-csv-demo
#| eval: false
# reading the data
# this should work when the file is in your working directory
ideal <- read_csv("ideal.csv")
```

## Reading an Excel File

Excel workbooks carry hidden complexity including multiple sheets, merged header rows, dates stored as serial numbers.

- `read_excel()` exposes each of these as an explicit argument, making your assumptions visible and reproducible.

```{r}
#| label: read-excel-demo
#| eval: false
# this reads the first sheet by default
ideal <- read_excel("ideal.xlsx")

# precise reading, when sheets have merged titles or extra header rows
ideal <- read_excel(
  "ideal.xlsx", # excel file
  sheet = "Calf Data", # specified sheet name
  skip = 2, # skip merged title rows before the header
  col_names = TRUE # use the first row as column names
)
```

#  The Dataset We'll Use for Today's Session

The **IDEAL study** followed 534 calves across 14 sub-locations in Western Kenya, recording health indicators at roughly six-week intervals from birth to one year.

## Loading the IDEAL Dataset

Three lines take you from nothing to a live dataset. In practice, you run these once at the top of every script.

```{webr-r}
#| label: show-data
# A tibble prints its dimensions and column types automatically
ideal
```

> Once a package is installed, you never need `install.packages()` again — only `library()`.

#  Inspecting Your Data

Never analyse data you have not looked at. These three functions are your first line of due diligence — run them immediately after every `read_*()` call.

## Preview

- `head()` shows the first 5 rows.
- `dim()` confirms the number of rows and columns.
- `names()` lists every column

:::: {.columns}
::: {.column width="50%"}
```{webr-r}
#| label: inspect-head
head(ideal, 5)
```
:::
::: {.column width="50%"}
```{webr-r}
#| label: inspect-dim
dim(ideal)
names(ideal)
```
:::
::::

## `glimpse()` — The Structural View

- `glimpse()` transposes your data: each column becomes a row, showing its type and first few values side by side.
- In a single screen, you learn the shape, the types, and the range of values.

```{webr-r}
#| label: inspect-glimpse
glimpse(ideal)
```

> Notice: `CalfSex` is an integer, not a label. Columns like `Theileria.spp.` contain dots in their names. `ReasonsLoss1` records the outcome. These are exactly the things we will fix with `rename()` and `mutate()`.

## `summary()` — The Statistical View

- `summary()` applies summary statistics to every numeric column. Missing values appear as a named count.
- One call brings out outliers, impossible values, and data completeness issues that would otherwise stay hidden until analysis.

```{webr-r}
#| label: inspect-summary
summary(ideal[, c("Weight", "ManualPCV", "Theileria.spp.", "ADWG")])
```

#  The Pipe Operator & `filter()`

The pipe is the single most transformative idea in tidyverse programming. It turns nested, inside-out function calls into a left-to-right sentence.

## The Pipe Operator `|>`

- Without the pipe, reading R code means parsing from the innermost call outward — the opposite of how you reason about a problem.
- The pipe lets you write steps in the order they happen.

:::: {.columns}
::: {.column width="70%"}
### Without the pipe

```{r}
#| label: no-pipe
#| eval: false
# read from the inside out — unnatural
select(
  filter(ideal, ReasonsLoss1 == "survived"),
  CalfID, sublocation, Weight
)
```
:::
::: {.column width="50%"}
### With the pipe

```{r}
#| label: with-pipe
#| eval: false
# read left to right — matches how you think
ideal |>
  dplyr::filter(ReasonsLoss1 == "survived") |>
  dplyr::select(CalfID, sublocation, Weight)
```
:::
::::

. . .

> Read `|>` as **"and then"** — *take the data, **and then** filter, **and then** select.* Keyboard shortcut: **Ctrl + Shift + M** in RStudio.

## `%>%` vs `|>` — Which Pipe?

| | `%>%` (magrittr) | `\|>` (base R)|
|--|------------------|----------------|
| **Requires a package?** | Yes — loaded with tidyverse | No — built into R ≥ 4.1 |
| **Speed** | Slightly slower | Marginally faster |
| **Where you'll see it** | Most existing tidyverse tutorials | Modern R code, including this session |

```{r}
#| label: pipe-both
#| eval: false
# these two lines produce identical output
ideal %>% filter(ReasonsLoss1 == "survived")
ideal |>  filter(ReasonsLoss1 == "survived")
```

## `filter()` — Keeping the Rows You Need

`filter()` evaluates a condition row by row and returns only the rows where the condition is `TRUE`. Every other row is silently discarded leaving the column structure is unchanged.

```{webr-r}
#| label: filter-basic
# keep only calves that did not survive
ideal |>
  filter(ReasonsLoss1 == "died")
```

## `filter()` — Combining Conditions

- Separate conditions with a `&` for AND logic, or `|` for OR logic.
- R tests each row independently — only rows that satisfy every condition are kept.

```{webr-r}
#| label: filter-and
# Theileria-positive calves that survived — AND logic
ideal |>
  dplyr::filter(`Theileria.spp.` == 1 & ReasonsLoss1 == "survived") |>
  head()
```

. . .

```{webr-r}
#| label: filter-or
# visits from either Kidera or Igero — OR logic
ideal |>
  dplyr::filter(sublocation == "Kidera" | sublocation == "Igero") |>
  head()
```

> Column names containing `.` require backticks: `` `Theileria.spp.` ``. This friction is precisely why `rename()` matters — which we cover shortly.

## `filter()` — Handling Missing Values

Missing values are R's most common source of silent, hard-to-detect errors. The rule is simple: `NA == NA` returns `NA`, not `TRUE`. That means `filter(Weight == NA)` always returns zero rows — even when missing values exist.

```{webr-r}
#| label: filter-na
# use is.na() to test for missingness
ideal |>
  filter(!is.na(Weight))
```

. . .

```{webr-r}
#| label: filter-na-check
# count how many visits are missing a weight measurement
ideal |>
  filter(is.na(Weight)) |>
  nrow()
```

> `is.na()` is the only reliable way to test for missing values in R.

## `filter()` — Reference: Logical Operators

| Operator | Meaning | Example |
|----------|---------|---------|
| `==` | Equal to | `ReasonsLoss1 == "survived"` |
| `!=` | Not equal to | `sublocation != "Igero"` |
| `>` / `<` | Greater / less than | `Weight > 30` |
| `>=` / `<=` | Greater / less or equal | `ManualPCV >= 24` |
| `&` | AND — both conditions true | `` `Theileria.spp.` == 1 & ReasonsLoss1 == "died"`` |
| `\|` | OR — either condition true | `sublocation == "Kidera" \| sublocation == "Igero"` |
| `%in%` | Belongs to a set | `sublocation %in% c("Kidera", "Igero", "Bukati")` |
| `is.na()` | Is missing | `filter(is.na(Weight))` |
| `between()` | Falls within a range | `between(Age, 100, 200)` |

## `filter()` — The `%in%` Operator

- `%in%` tests membership in a set — far more readable than chaining multiple `==` conditions with `|`.
- It scales naturally: adding a fourth location is one comma, not a new line of logic.

```{webr-r}
#| label: filter-in
# select three specific sub-locations
ideal |>
  dplyr::filter(sublocation %in% c("Kidera", "Igero", "Luanda")) |>
  dplyr::select(CalfID, sublocation, Weight, ReasonsLoss1) |>
  head()
```

# ️ `select()` & `rename()`

- A well-structured analysis uses only the columns it needs, named in a way that makes code self-explanatory.
- `select()` shapes the column space;
- `rename()` removes the friction of working with raw source names.

## `select()` — Choosing Your Columns

Pass column names directly. `select()` returns a tibble with exactly those columns in the order you listed them. Rows are untouched.

```{webr-r}
#| label: select-basic
ideal |>
  dplyr::select(CalfID, sublocation, CalfSex, Weight, ManualPCV, `Theileria.spp.`, ReasonsLoss1)
```

> Columns with `.` in their names still require backticks inside `select()`. One of many reasons to `rename()` early.

## `select()` — Dropping Columns

Prefix any name with `-` to drop it instead of keeping it. When you want most columns but not a few, dropping is more concise than listing every keeper.

```{webr-r}
#| label: select-drop
# removing columns
ideal |>
  select(-CADOB, -VisitID, -VisitWeek, -WeightMax, -WeightMin)
```

## `select()` — Helper Functions

When a dataset has dozens of columns, selecting by pattern is more robust than listing names.

```{webr-r}
#| label: select-helpers
# select both ELISA columns without naming them explicitly
ideal |>
  select(CalfID, starts_with("ELISA")) |>
  head(n = 1)
```

| Helper | Selects columns whose names… |
|--------|------------------------------|
| `starts_with("x")` | Begin with "x" |
| `ends_with("x")` | End with "x" |
| `contains("x")` | Contain "x" anywhere |
| `where(is.numeric)` | Pass a type test |
| `everything()` | Are not already selected |

## `rename()`

- You have now seen the friction that raw source names create — backticks for `Theileria.spp.`, uppercase everywhere, cryptic codes like `ReasonsLoss1`. `rename()` removes that friction permanently. Do it once, and every line that follows is cleaner.

```{webr-r}
#| label: rename-basic
# syntax: rename(new_name = old_name)
ideal |>
  rename(
    sex       = CalfSex,
    theileria = `Theileria.spp.`,
    outcome   = ReasonsLoss1
  ) |>
  select(CalfID, sex, theileria, outcome) |>
  head(2)
```

## `arrange()` — Sorting Rows

- `arrange()` reorders rows by one or more columns. Unlike `filter()`, it never removes rows it only changes their order.

```{webr-r}
#| label: arrange
# which calves were heaviest at the time of their visit?
ideal |>
  filter(!is.na(Weight)) |>
  select(CalfID, sublocation, Age, Weight, ReasonsLoss1) |>
  arrange(desc(Weight)) |> # from largest to smallest.
  head(5)
```

#  `mutate()`

- Raw data rarely arrives in the form your analysis needs.
- `mutate()` derives new columns from existing ones without altering a single original value. Think of it as adding new knowledge to your data, not replacing old knowledge.

## `mutate()` — Creating a New Column

- `mutate()` appends the new column to the right of the tibble. Every existing column is preserved exactly as it was.

```{webr-r}
#| label: mutate-basic
# here we decode the integer sex code into a human-readable label.
ideal |>
  mutate(sex_label = if_else(CalfSex == 1, "Male", "Female")) |>
  select(CalfID, CalfSex, sex_label, Weight, ReasonsLoss1) |>
  head(5)
```

## `mutate()` — Multiple Columns at Once

- A single `mutate()` call creates as many columns as you need. Later expressions in the same call can reference columns created earlier, so the order of your derivations matters and can build on itself.

```{webr-r}
#| label: mutate-multi
ideal |>
  mutate(
    sex_label  = if_else(CalfSex == 1, "Male", "Female"),
    weight_lbs = round(Weight * 2.205, 1), # convert kg to pounds
    anaemic = ManualPCV < 24 # confirmatory test for anaemia
  ) |>
  select(CalfID, sex_label, Weight, weight_lbs, ManualPCV, anaemic) |>
  head(3)
```

## `mutate()` with `case_when()`

- `case_when()` is R's vectorised multi-condition classifier — the equivalent of Excel's nested `IFS`.
- Each condition is evaluated top-to-bottom; the first `TRUE` match determines the value assigned to that row.

```{webr-r}
#| label: case-when
ideal |>
  dplyr::mutate(
    anaemia_class = case_when(
      is.na(ManualPCV)  ~ "Unknown",       # always test for NA first
      ManualPCV < 24    ~ "Severe anaemia",
      ManualPCV < 30    ~ "Mild anaemia",
      TRUE ~ "Normal" 
    )
  ) |>
  dplyr::select(CalfID, ManualPCV, anaemia_class)
```

## `case_when()` — The Mental Model

Each line is one rule: **"When** *this condition* **is true → assign** *this value*."

```{r}
#| label: case-when-structure
#| eval: false
mutate(
  new_column = case_when(
    condition_1  ~ "value when condition_1 is TRUE",
    condition_2  ~ "value when condition_2 is TRUE",
    condition_3  ~ "value when condition_3 is TRUE",
    TRUE ~ "value for everything else"
  )
)
```

Rules are tested in order. The moment a condition matches, R stops testing — so place the most specific conditions first, the broadest last.

#  `group_by()` & `summarise()`

Most research questions are comparisons: *does this outcome differ across groups?* `group_by()` and `summarise()` are how tidyverse answers them

## `summarise()` — The Baseline

Without a grouping, `summarise()` collapses the entire dataset to a single summary row. This is your reference point.

```{webr-r}
#| label: summarise-alone
ideal |>
  summarise(
    n_visits = n(),
    n_calves = n_distinct(CalfID),
    mean_weight = round(mean(Weight,    na.rm = TRUE), 1),
    mean_pcv = round(mean(ManualPCV, na.rm = TRUE), 1),
    theileria_pos = sum(`Theileria.spp.` == 1, na.rm = TRUE)
  )
```

> `na.rm = TRUE` tells R to ignore missing values when computing the mean. Without it, any column containing a single `NA` returns `NA` as the summary.

## `group_by()` + `summarise()`

- Add `group_by()` and every summary is computed independently for each group. The result is one row per unique combination of grouping variables.
- Call `ungroup()` afterwards to release the grouping structure.

```{webr-r}
#| label: group-summarise
result <- ideal |>
  group_by(sublocation, ReasonsLoss1) |>
  summarise(
    n_calves = n_distinct(CalfID),
    mean_weight = round(mean(Weight,    na.rm = TRUE), 1),
    mean_pcv = round(mean(ManualPCV, na.rm = TRUE), 1)
  ) |>
  ungroup() |>
  arrange(sublocation, ReasonsLoss1)
result
```

## Disease Burden by Outcome

```{webr-r}
#| label: group-theileria
ideal |>
  group_by(ReasonsLoss1) |>
  summarise(
    n_calves = n_distinct(CalfID),
    theileria_pos = sum(`Theileria.spp.` == 1, na.rm = TRUE),
    theileria_pct = round(mean(`Theileria.spp.` == 1, na.rm = TRUE) * 100, 1),
    mean_pcv = round(mean(ManualPCV, na.rm = TRUE), 1),
    mean_adwg = round(mean(ADWG,      na.rm = TRUE), 3)
  ) |>
  ungroup()
```

## Common Summary Functions

| Function | Computes |
|----------|----------|
| `n()` | Number of rows in the group |
| `n_distinct(x)` | Number of unique values |
| `sum(x, na.rm = TRUE)` | Total — events, counts, doses |
| `mean(x, na.rm = TRUE)` | Average |
| `median(x, na.rm = TRUE)` | Median — robust to outliers |
| `sd(x, na.rm = TRUE)` | Standard deviation |
| `min(x)` / `max(x)` | Minimum and maximum |
| `sum(is.na(x))` | Count of missing values |
| `mean(condition)` | Proportion satisfying a condition |

> Every numeric summary on real data should carry `na.rm = TRUE`. Missing values will otherwise silently propagate as `NA` in your output.

#  Joins

Real datasets live across multiple tables. Joins are how you reunite them — matching rows from two sources on a shared key, without manually copying values between files.

## Why Joins?

The IDEAL dataset records one row per calf visit. A separate table holds sub-location metadata — division, altitude, administrative boundaries. Joining enriches every visit row without duplicating the metadata by hand.

:::: {.columns}
::: {.column width="50%"}
**`ideal`** — Visit records

```{r}
#| label: join-left-preview
#| echo: false
ideal |> distinct(sublocation) |> arrange(sublocation)
```
:::
::: {.column width="50%"}
**`sublocation_info`** — Area metadata

```{r}
#| label: join-right-preview
#| echo: false
sublocation_info
```
:::
::::

## The Four Join Types

| Join | Result |
|------|--------|
| `left_join(x, y)` | All rows from `x`; matched columns from `y`; unmatched rows get `NA` |
| `right_join(x, y)` | All rows from `y`; matched columns from `x` |
| `inner_join(x, y)` | Only rows with a match in both tables |
| `full_join(x, y)` | All rows from both; unmatched cells get `NA` |

> `left_join()` is the workhorse of data analysis. It preserves your primary dataset intact and attaches reference information wherever a match exists. Unmatched rows are retained with `NA`, not silently lost.

## `left_join()` — Attach Without Losing Rows

```{webr-r}
#| label: left-join
# summarise to sublocation level, then attach geographic metadata
calf_summary <- ideal |>
  group_by(sublocation) |>
  summarise(
    n_calves = n_distinct(CalfID),
    mean_pcv = round(mean(ManualPCV, na.rm = TRUE), 1)
  ) |>
  ungroup()

calf_join <- calf_summary |>
  left_join(sublocation_info, by = "sublocation")

calf_join
```

> **Yiro West** has no entry in `sublocation_info`, so it receives `NA` for `division` and `altitude_m`. It is still present in the result — `left_join()` never drops a row from the left table.

## `inner_join()` — Only Complete Matches

```{webr-r}
#| label: inner-join
# only sub-locations present in both tables survive
calf_summary |>
  inner_join(sublocation_info, by = "sublocation")
```

> **Yiro West** is now absent — it had no match on the right. Use `inner_join()` when your analysis requires complete information for every row, and you are prepared to lose unmatched records.

## `anti_join()` — Diagnosing the Gaps

```{webr-r}
#| label: anti-join
# which sub-locations in the data have no metadata entry?
calf_summary |>
  anti_join(sublocation_info, by = "sublocation")
```

> `anti_join()` returns only the rows that *failed* to match. It is a data quality diagnostic. Run it before any analysis that depends on a join to catch missing reference records.

## Joining on the Full Dataset

```{webr-r}
#| label: join-raw
ideal |>
  left_join(sublocation_info, by = "sublocation") |>
  filter(!is.na(Weight)) |>
  select(CalfID, sublocation, division, altitude_m, Weight, ReasonsLoss1) |>
  head(8)
```

> When you join on the full visit-level dataset, each metadata value repeats for every row belonging to that sub-location. This is correct — it is how tidy, rectangular data works.

#  Data Visualisation in R

`ggplot2` translates data frames into graphics through three composable layers that stack with `+`: **data**, **aesthetics** (`aes()`), and **geometry** (`geom_*()`). Everything else — themes, labels, scales — is optional polish on top.

## `geom_bar()` — Counting Outcomes {.code-only}

```{webr-r}
#| label: bar-plot
ggplot(data    = ideal,
       mapping = aes(x = ReasonsLoss1)) +
  geom_bar(fill = "#185FA5") +
  theme_classic() +
  labs(title = "Survival of calves in the first year of life",
       x = "",
       y = "Number of visits")
```


## `geom_bar()` — Result

:::: {.columns}
::: {.column width="60%"}
```{r}
#| label: bar-plot-static
#| echo: false
#| fig-width: 5
#| fig-height: 3.5
#| out-width: "100%"
ggplot(data    = ideal,
       mapping = aes(x = ReasonsLoss1)) +
  geom_bar(fill = "#185FA5") +
  theme_classic() +
  labs(title = "Survival of calves in the first year of life",
       x = "",
       y = "Number of visits")
```
:::
::: {.column width="40%"}
`geom_bar()` counts rows for you — supply only `x` and it derives bar heights automatically.

**What to notice**

- Two categories: `died`, `survived`
- `survived` is the largest group by a clear margin
- y-axis counts **visits**, not unique calves — one calf appears at every visit

> **Try it:** go back and change `fill = "#185FA5"` to `fill = "#639922"`, re-run, then advance again to see the updated chart.
:::
::::

## Preparing the Data — `calves_per_sub`

Before using `geom_col()`, summarise to one row per sub-location. 

```{webr-r}
#| label: col-data-prep
calves_per_sub <- ideal |>
  group_by(sublocation) |>
  summarise(n_calves = n_distinct(CalfID)) |>  # unique calves
  ungroup()

calves_per_sub
```

> `n_distinct(CalfID)` counts unique calves per sub-location. Notice the spread — some sub-locations recruited nearly twice as many calves as others.

## `geom_col()` — Pre-summarised Data {.code-only}

```{webr-r}
#| label: col-basic
ggplot(data = calves_per_sub,
       mapping = aes(x = sublocation, y = n_calves)) +
  geom_col(fill = "#185FA5") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "", y = "Number of calves",
       title = "Calves recruited per sub-location")
```


## `geom_col()` — Result

:::: {.columns}
::: {.column width="60%"}
```{r}
#| label: col-basic-static
#| echo: false
#| fig-width: 5.5
#| fig-height: 4.5
#| out-width: "100%"
calves_per_sub <- ideal |>
  group_by(sublocation) |>
  summarise(n_calves = n_distinct(CalfID)) |>
  ungroup()

ggplot(calves_per_sub,
       aes(x = sublocation, y = n_calves)) +
  geom_col(fill = "#185FA5") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "", y = "Number of calves",
       title = "Calves recruited per sub-location")
```
:::
::: {.column width="40%"}
`geom_col()` requires both `x` and `y` — you supply the heights you already computed.

| | `geom_bar()` | `geom_col()` |
|--|--|--|
| You supply | `x` only | `x` and `y` |
| R computes | counts | nothing |
| Use when | raw data | summarised data |

Notice the overlapping x-axis labels — a common problem with many categories. The next slide shows the fix.

> **Try it:** go back and add `fill = sublocation` inside `aes()` to colour each bar differently, then advance again.
:::
::::

## `coord_flip()` and `reorder()` {.code-only}

```{webr-r}
#| label: col-flipped
ggplot(data = calves_per_sub,
       mapping = aes(x = reorder(sublocation, n_calves),
                     y = n_calves)) +
  geom_col(fill = "#185FA5") +
  coord_flip() +
  theme_classic() +
  labs(x = "",
       y = "Number of calves",
       title = "Calves recruited per sub-location")
```


## `coord_flip()` — Result

:::: {.columns}
::: {.column width="60%"}
```{r}
#| label: col-flipped-static
#| echo: false
#| fig-width: 5.5
#| fig-height: 4.5
#| out-width: "100%"
ggplot(calves_per_sub,
       aes(x = reorder(sublocation, n_calves), y = n_calves)) +
  geom_col(fill = "#185FA5") +
  coord_flip() +
  theme_classic() +
  labs(x = "",
       y = "Number of calves",
       title = "Calves recruited per sub-location")
```
:::
::: {.column width="40%"}
This transform the chart into a ranked, readable comparison.

- **`reorder(sublocation, n_calves)`** — sorts bars by value instead of alphabetically; the longest bar is always at top
- **`coord_flip()`** — rotates axes so long labels read horizontally without overlapping

> **Try it:** go back and change `reorder(sublocation, n_calves)` to `reorder(sublocation, -n_calves)` to reverse the sort order, then advance again.
:::
::::

## Preparing the Data — `calves_by_sex`

To split bars by gender , we need two grouping variables. Run this before the next slide.

```{webr-r}
#| label: sex-data-prep
calves_by_sex <- ideal |>
  mutate(sex_label = case_when(
    CalfSex == 1 ~ "Male",
    CalfSex == 2 ~ "Female",
    TRUE ~ "Unknown"
  )) |>
  group_by(sublocation, sex_label) |>
  summarise(n_calves = n_distinct(CalfID), .groups = "drop")

calves_by_sex
```

> The table now has **two rows per sub-location** — one for each sex. The next slide maps `CalfSex` to `fill` to split the bars.

## Grouped Bars — the `fill` Aesthetic {.code-only}

```{webr-r}
#| label: bar-grouped
top6_subs <- calves_by_sex |>
  group_by(sublocation) |>
  summarise(total = sum(n_calves)) |>
  slice_max(total, n = 5) |>
  pull(sublocation)

ggplot(filter(calves_by_sex, sublocation %in% top6_subs),
       aes(x = reorder(sublocation, n_calves),
           y = n_calves, fill = sex_label)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(Male = "#2196F3", Female = "#E91E63")) +
  coord_flip() + theme_classic() +
  labs(x = "", y = "Number of calves", fill = "Sex",
       title = "Calves per sub-location by sex (top 5)")
```


## Grouped Bars — Result

:::: {.columns}
::: {.column width="60%"}
```{r}
#| label: bar-grouped-static
#| echo: false
#| fig-width: 5.5
#| fig-height: 4
#| out-width: "100%"
calves_by_sex <- ideal |>
  mutate(sex_label = case_when(
    CalfSex == 1 ~ "Male",
    CalfSex == 2 ~ "Female",
    TRUE ~ "Unknown"
  )) |>
  group_by(sublocation, sex_label) |>
  summarise(n_calves = n_distinct(CalfID), .groups = "drop")

top6_subs <- calves_by_sex |>
  group_by(sublocation) |>
  summarise(total = sum(n_calves)) |>
  slice_max(total, n = 5) |>
  pull(sublocation)

ggplot(filter(calves_by_sex, sublocation %in% top6_subs),
       aes(x = reorder(sublocation, n_calves),
           y = n_calves, fill = sex_label)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(Male = "#2196F3", Female = "#E91E63")) +
  coord_flip() + theme_classic() +
  labs(x = "", y = "Number of calves", fill = "Sex",
       title = "Calves per sub-location by Gender  (top 5)")
```
:::
::: {.column width="40%"}
Mapping `fill = sex_label` tells ggplot2 to split each bar by sex and colour the groups automatically. `scale_fill_manual()` replaces the defaults with intentional colours.

**`position` options**

| Value | Effect |
|-------|--------|
| `"dodge"` | Side by side — compare values |
| `"stack"` | Stacked — compare totals |
| `"fill"` | Stacked to 100% — compare proportions |

> **Try it:** go back, remove the `top6_subs` filter and use `calves_by_sex` directly, switch `position` to `"stack"`, then advance again.
:::
::::

#  Mapping in R

## Maps in R — `sf` and `ggplot2`

Spatial analysis in R centres on the **`sf`** package — *simple features* — which stores geographic shapes as ordinary data frames with a special **geometry column** attached. Every dplyr verb works on `sf` objects: `filter()`, `mutate()`, `left_join()`, and so on.

Two key ideas to remember:

- `read_sf()` reads a GeoPackage (or shapefile) into a data frame with geometry — exactly like `read_csv()` reads a CSV
- `geom_sf()` renders the geometry column automatically inside ggplot2 — no coordinate wrangling needed

> The `kenya` object is already loaded in this session — downloaded from GitHub the same way as `ideal3a.csv`. Inspect it below:

```{webr-r}
#| label: sf-inspect
kenya
```

## `geom_sf()` — Kenya Boundaries {.code-only}

```{webr-r}
#| label: kenya-outline
ggplot(data = kenya) +
  geom_sf(fill = "white", colour = "black", linewidth = 0.4) +
  theme_void() +
  labs(title = "Kenya — County Boundaries (47 Counties)")
```


## `geom_sf()` — Result

:::: {.columns}
::: {.column width="60%"}
```{r}
#| label: kenya-outline-static
#| echo: false
#| fig-width: 5
#| fig-height: 4.5
#| out-width: "100%"
ggplot(data = kenya) +
  geom_sf(fill = "white", colour = "black", linewidth = 0.4) +
  theme_void() +
  labs(title = "Kenya — County Boundaries (47 Counties)")
```
:::
::: {.column width="40%"}
`geom_sf()` is the spatial equivalent of `geom_col()` — you supply a data frame and ggplot2 handles all coordinate projection and border drawing.

**What you get**

- All 47 county borders from a single function call
- The `geometry` column is detected automatically — no `x =` or `y =` needed

> **Try it:** change `fill = "#f0f4f8"` to `fill = "#185FA5"` and re-run to colour every county blue.
:::
::::

## Preparing the Choropleth — Cattle Population

Filter the population data to one species and build the joined spatial table.

```{webr-r}
#| label: choropleth-prep
# filter to a single species
cattle <- population |>
  filter(Species == "Cattle(Total)")

# county names in the shapefile don't always match the CSV exactly —
# case_when() fixes four known mismatches before the join
cattle_map <- kenya |>
  rename(county = ADM1_EN) |>
  mutate(county = case_when(
    county == "Murang'a" ~ "Muranga",
    county == "Elgeyo-Marakwet" ~ "Elgeyo Marakwet",
    county == "Tharaka-Nithi" ~ "Tharaka Nithi",
    county == "Trans Nzoia" ~ "Trans-Nzoia",
    TRUE ~ county
  )) |>
  left_join(cattle, by = "county")

cattle_map |> select(county, value)
```

> The `case_when()` step is a real-world data quality step — spelling mismatches between datasets are extremely common. Always check with `setdiff(table_a$key, table_b$key)` before any join.

## Choropleth — Cattle Population {.code-only}

```{webr-r}
#| label: choropleth-plot
ggplot(data = cattle_map) +
  geom_sf(aes(fill = value), colour = "black", linewidth = 0.3) +
  scale_fill_gradient(low = 'lightyellow', 
                      high = 'darkred', 
                      labels = scales::comma,
                      name = "Cattle") +
  theme_void() +
  labs(title = "Cattle Population by County — Kenya")
```


## Choropleth — Result

:::: {.columns}
::: {.column width="60%"}
```{r}
#| label: choropleth-static
#| echo: false
#| fig-width: 5
#| fig-height: 4.5
#| out-width: "100%"
cattle <- population |>
  filter(Species == "Cattle(Total)")
cattle_map <- kenya |>
  rename(county = ADM1_EN) |>
  mutate(county = case_when(
    county == "Murang'a" ~ "Muranga",
    county == "Elgeyo-Marakwet" ~ "Elgeyo Marakwet",
    county == "Tharaka-Nithi" ~ "Tharaka Nithi",
    county == "Trans Nzoia" ~ "Trans-Nzoia",
    TRUE ~ county
  )) |>
  left_join(cattle, by = "county")
ggplot(data = cattle_map) +
  geom_sf(aes(fill = value), colour = "black", linewidth = 0.3) +
  scale_fill_gradient(low = 'lightyellow', 
                      high = 'darkred', 
                      labels = scales::comma,
                      name = "Cattle") +
  theme_void() +
  labs(title = "Cattle Population by County — Kenya")
```
:::
::: {.column width="40%"}
`left_join()` connects the spatial data frame to the CSV table on the shared `county` key — ggplot2 maps `value` to `fill` exactly as it would for a bar chart.

**Reading the map**

- Darker red = larger cattle population
- Turkana county typically show the highest density

> **Try it:** go back, change `"Cattle(Total)"` to `"Goats"` in the `filter()`, re-run both slides, then advance — observe how the spatial pattern shifts between species.
:::
::::

## ggplot2 Reference — Common Geometries

| Geometry | Best for | Key aesthetics |
|----------|----------|----------------|
| `geom_bar()` | Counts from raw data | `x`, `fill` |
| `geom_col()` | Pre-summarised counts / values | `x`, `y`, `fill` |
| `geom_point()` | Scatter — two continuous variables | `x`, `y`, `colour`, `size` |
| `geom_line()` | Trends over time | `x`, `y`, `group`, `colour` |
| `geom_boxplot()` | Distribution by group | `x`, `y`, `fill` |
| `geom_histogram()` | Distribution of one variable | `x`, `fill` |
| `geom_sf()` | Spatial features / maps | `fill`, `colour` |

> Every geometry inherits aesthetics from `ggplot(aes(...))`. You override them inside the `geom_*()` call when one layer needs different mappings.